---
title: "Section 21 Plotting with R (II): making maps"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

***

# Prerequisites

Load the libraries with R:

```{r}
library(rgdal)
library(ggplot2)
library(plyr)
library(dplyr)
library(ggsn)
library(raster)
library(ggmap)
library(mapproj)
```

***

# Shape file

*ESRI shapefile* (or shape file) is widely used in environmental science and geoscience. A shapefile consists of various files of the same name, but with different extensions. They should all be in one directory. Here we use the shape file of China map as an example, [download it](https://zhu-group.github.io/ese335/download/China_map.zip). Decompress it. Open the folder, you will see there are three files of different formats (`.dbf`, `.shp`, `.shx`). In R, we can read the shape file via:

```{r}

# Location of the .shp file 
Local_path <- "China_map/"

# Read china map, a shape file
China_map <- rgdal::readOGR(paste0(Local_path, "bou2_4p.shp"))
```

```{r}

# Check the attributes 
class(China_map)
# Show variable names
names(China_map)
```

From here, we can see that there are `925` polygons, and R just follows the polygons defined by this object then line them each one up. If you want to see the data hidden in `SpatialPolygonsDataFrame`, we can use the operator `@`:

```{r}
# Check the attributes, use the operator @
map_data <- China_map@data
head(map_data)
tail(map_data)
```

```{r}
# Convert characters
China_map$NAME <- iconv(China_map$NAME, "GBK")
# Check the attributes, use the operator @
map_data <- China_map@data
head(map_data)
tail(map_data)
```

Use `ggplot()` to plot the shape file:

```{r}

# Quick plot use ggplot
ggplot(China_map, aes(x = long, y = lat, group = group)) +
  geom_path(color = "grey40") +
  geom_polygon(fill = 'lightblue')
```

This is certainly not very good-looking. Let's first assign each province with an id:

```{r}

# Assign each province with an ID
China_map2     <- data.frame(China_map, id=seq(0:924)-1) 

# Join two data sets by NAME
China_map_new  <- join(fortify(China_map), China_map2, type = "full")
```

Here we use the `fortify()` function to include the lat and lon information. Now the new shape file can be plotted as:

```{r}
# Plot the new shape file
ggplot(China_map_new, aes(x = long, y = lat, group = id, fill = NAME)) +
  # Plot the border
  geom_path(color = 'grey40') +
  geom_polygon() +
  # Using different colors
  scale_fill_manual(values = rainbow(33), guide = F) +
  coord_map()
```

We can also edit the original shape file by adding new information. For example, here we want to add both `NAME_EN` and `POP` onto the shape file:

```{r}

# Name in Chinese
# This is the same as the NAME in the original shape file
NAME    <- c("北京市"  , "天津市"   , "河北省"      , "山西省"   , "内蒙古自治区"    , 
             "辽宁省"  , "吉林省"   , "黑龙江省"    , "上海市"   , "江苏省"          , 
             "浙江省"  , "安徽省"   , "福建省"      , "江西省"   , "山东省"          , 
             "河南省"  , "湖北省"   , "湖南省"      , "广东省"   , "广西壮族自治区"  , 
             "海南省"  , "重庆市"   , "四川省"      , "贵州省"   , "云南省"          , 
             "西藏自治区", "陕西省" , "甘肃省"      , "青海省"   , "宁夏回族自治区"  , 
             "新疆维吾尔自治区"     , "台湾省"      , "香港特别行政区")

# Name in English accordingly
NAME_EN <- c("Beijing" , "Tianjin"  , "Hebei"       , "Shanxi"   , "Inner Mongolia"  ,
             "Liaoning", "Jilin"    , "Heilongjiang", "Shanghai" , "Jiangsu"         ,
             "Zhejiang", "Anhui"    , "Fujian"      , "Jiangxi"  , "Shandong"        ,
             "Henan"   , "Hubei"    , "Hunan"       , "Guangdong", "Guangxi"         ,
             "Hainan"  , "Chongqing", "Sichuan"     , "Guizhou"  , "Yunnan"          ,
             "Tibet"   , "Shaanxi"  , "Gansu"       , "Qinghai"  , "Ningxia"         ,
             "Xinjiang"             , "Taiwan"      , "Hong Kong"     )

# Population of each province, from the 2021 census
Pop     <- c(21893095  , 13866009   , 74610235      ,  34915616  ,  24049155         , 
             42591407  , 24073453   , 31850088      ,  24870895  ,  84748016         , 
             64567588  , 61027171   , 41540086      ,  45188635  , 101527453         , 
             99365519  , 57752557   , 66444864      , 126012510  ,  50126804         , 
             10081232  , 32054159   , 83674866      ,  38562148  ,  47209277         , 
              3648100  , 39528999   , 25019831      ,   5923957  ,   7202654         , 
             25852345               , 23561236      ,   7474200       )

# Make data frame
Popdata <- data.frame(NAME, NAME_EN, Pop)

# Joint the data by NAME
China_map_pop <- join(China_map_new, Popdata, type = "full")

# Plot
ggplot(China_map_pop, aes(x = long, y = lat, group = id, fill = Pop)) +
  geom_polygon() +
  geom_path(color = "grey40") +
  coord_map()
```

With some efforts, this can plot can be made much better:

```{r}
ggplot(China_map_pop, aes(x = long, y = lat, group = group, fill=Pop)) +
  labs(fill = "Population")+
  geom_polygon()+
  geom_path(color = "grey40")+
  # Try a different color theme
  scale_fill_gradientn(colours=rev(heat.colors(20)),na.value="grey90",
                       guide = guide_colourbar(barwidth = 0.8, barheight = 10)) + 
  # Projects a portion of the earth
  coord_map() + 
  # Change theme
  theme_classic() + 
  # Change labels
  labs(fill = "Population", x = "Longitude", y = "Latitude") + 
  # Add map scale
  ggsn::scalebar(data = China_map_pop, dist = 500, dist_unit = "km",
                 border.size = 0.4, st.size = 2,
                 box.fill = c('black','white'),
                 transform = TRUE, model = "WGS84") 
```

You can only select one (or more) province and add one point to the plot:

```{r}

## Extract Guangdong and Hong Kong from the data
GD_HK <- subset(China_map_pop, NAME == "广东省" | NAME == "香港特别行政区") 

# Plot
ggplot(GD_HK,aes(x = long, y = lat, group = id)) +
  # Polygon
  geom_polygon(fill = "lightblue") +
  # Border
  geom_path(color = "lightblue") +
  # Add one point: SUSTech (lat:22.59670, lon:113.98201)
  geom_point(x = 113.98201, y = 22.59670, fill = NA) +
  # Add label
  annotate("text", x = 113.98201, y = 22.59670+0.25, label = "SUSTech") +
  # Projects a portion of the earth
  coord_map()
```

***

# Raster data

**Raster** files have a much more compact data structure than vectors. Because of their regular structure, the coordinates do not need to be recorded for each pixel or cell in the rectangular extent. A raster is defined by:  

* CRS (coordinate reference system)

* Coordinates of its origin

* A distance or cell size in each direction

* A dimension or numbers of cells in each direction

* An array of cell values

Given this structure, coordinates for any cell can be computed and don’t need to be stored.

The `raster` package is a major extension of spatial data classes to access large rasters and in particular, to process very large files. It includes `RasterLayer` and functions for converting among different classes and operators for computations on the raster data.

We can use `raster()` to read in raster files. Download this [tiff file](https://zhu-group.github.io/ese335/download/wc2.1_10m_wind_11.tif) (`wc2.1_10m_wind_11.tif`, ~ 2.4MB), which is the mean wind speed (m s^-1^) in Nov. from 1970-2000 provided by the [WorldClim](https://worldclim.org/data/worldclim21.html). The data is at a resolution of `10` minutes (~ `340` km^2^).

In R, we can load it as:

```{r}

# Read the tiff file
Wind_Nov <- raster("wc2.1_10m_wind_11.tif")

# Look at the raster attributes
Wind_Nov
```

We can plot it using `plot()` or `image()` function:

```{r}

# Quick plot using plot()
plot(Wind_Nov, main="Wind speed in Nov.")

# Set color
col <- terrain.colors(30)

# Quick using image()
image(Wind_Nov,  main="Wind speed in Nov.", col=col)

# Add contour lines
contour(Wind_Nov, add=T, col="red")
```

Or with `ggplot()`, we can make a better map:

```{r}

# Convert the raster to a date.frame
Wind_Nov_df <- as.data.frame(Wind_Nov, xy = TRUE) 

# Check the data structure
str(Wind_Nov_df)

# Making plot
ggplot() +
  geom_raster(data = Wind_Nov_df,  
              aes(x = x, y = y, fill = wc2.1_10m_wind_11)) + 
  # Change labels
  labs(x="Latitude", y="Longitude") +
  # Change theme
  theme_bw() +
  coord_equal() +
  # Change legend
  scale_fill_gradient( "Wind speed (m/s)", limits=c(0,20),
                       low = "white",
                       high = "red",
                       n.breaks = 10,
                       space = "Lab",
                       na.value = "grey50",
                       guide = "colourbar",
                       aesthetics = "fill") +
  # Adjust the theme
  theme(axis.title.x = element_text(size=16),
        axis.title.y = element_text(size=16, angle=90),
        axis.text.x = element_text(size=14),
        axis.text.y = element_text(size=14),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "right",
        legend.key = element_blank()
  ) +
  # Add a title
  ggtitle("Wind Speed in Nov.") 
```

You can also use `crop()` function to select an area of interest:  

```{r}

# Define the crop extent
Crop_box <- c(100,125,15,30)

# Crop the raster
Wind_Nov_crop <- crop(Wind_Nov, Crop_box)

# Plot cropped DEM
plot(Wind_Nov_crop, main="Wind speed in Nov.")

```

***

# Adding features to a base map

## Add points

Suppose we want to plot site locations on a city scale, and we want to add a base map that shows road, terrain, and built-up areas. This can be done with the `ggmap()` package:

```{r, eval=T}

# Site information
Site_name <- c("SUSTech", "Longhua", "Xichong", "Baoan", "Kuiyong")
Site_lon  <- c(114.06667,114.02200,114.56111,113.89606,114.42824)
Site_lat  <- c(22.61667,22.72882,22.48077,22.53965,22.63427)
Site_type <- c("Urban", "Urban", "Background", "Urban", "Rural")
  
# Make a data frame
Site_data <- data.frame(name=Site_name, lon=Site_lon, lat=Site_lat, type=Site_type)

# Get the lat and lon range
Mapbox    <- make_bbox(lon = Site_data$lon, lat = Site_data$lat, f = .1)

# Pull the base map
# The keyword zoom defines the map resolution
# Here we use base map from stadiamaps
# To obtain an API key and enable services, 
# go to https://client.stadiamaps.com/signup/
# Then use register_stadiamaps() to set the API key
Base_map  <- get_stadiamap(Mapbox, zoom = 10, maptype = "stamen_terrain")
ggmap(Base_map)

# Plot
ggmap(Base_map) + 
  # Add sites
  geom_point(data=Site_data, aes(x=lon, y=lat, fill=type, shape=type), 
             color="white", cex=5.5) + # plot the points
  # Change color
  scale_fill_manual(values = c("green", "blue", "red"), 
                    labels=c("Background", "Rural","Urban"), name=NULL) + 
  # Change shape
  scale_shape_manual(values = c(21,22,24), 
                     labels=c("Background", "Rural","Urban"), name=NULL) + 
  # Change labels and title
  labs(x="Latitude", y="Longitude", title="Monitoring sites") + # label the axes
  # Change theme
  theme_bw() + 
  theme(legend.position="bottom", 
        legend.key  = element_rect(colour = "white"), 
        axis.text   = element_text(size = rel(0.75)), 
        axis.text.x = element_text(angle=45, vjust=0.5)) 
```


## Add a path

Download the [track of Typhoon Mangkhut](https://zhu-group.github.io/ese335/download/Mangkhut.txt), and read via:

```{r}

# Read the data
Mangkhut_data       <- read.table("Mangkhut.txt")

# Get path and max wind speed
Mangkhut_lat        <- Mangkhut_data$V4*0.1
Mangkhut_lon        <- Mangkhut_data$V5*0.1
Mangkhut_pressure   <- Mangkhut_data$V6
Mangkhut_max_speed  <- Mangkhut_data$V7*0.514
```

let's plot the max wind speed on the map:

```{r, eval=T}

# Make a data frame for ggplot
Mangkhut_data_new <- data.frame(lat=Mangkhut_lat,lon=Mangkhut_lon,
                                pressure=Mangkhut_pressure, speed=Mangkhut_max_speed)

# Get the domain
Domain            <- make_bbox(lon = Mangkhut_data_new$lon, 
                               lat = Mangkhut_data_new$lat, f = .1)

# Get the base map
Base_map          <- get_stadiamap(Domain, zoom = 5, maptype = "stamen_terrain")

# Show base map
ggmap(Base_map)

# Plot base map
Map1              <- ggmap(Base_map)

# Plot the path
Map1 + 
  # Plot the track 
  geom_path(data = Mangkhut_data_new, 
            aes(color=speed), size=1.5, 
            lineend = "round") + 
  # Set the color
  scale_colour_gradient("Max Wind Speed (m/s)", low = "white", high = "red", 
                        breaks = seq(10, 60, by = 10)) +
  # Change labels
  labs(x="Latitude", y="Longitude", title="Typhoon Mangkhut Track") + # label the axes
  theme_bw() 
```


***

# In-class exercises

## Exercise #1

Plot wind speed in Nov. near your hometown (a `10` degree by `10` degree domain) using the `Wind_Nov` object.

## Exercise #2

Plot a map of your home city (zoom=12), and add a few points to the map.

***

# Further reading

* [Introduction to Spatial Data Types in R](https://cengel.github.io/rspatial/2_spDataTypes.nb.html)

* [Introduction to Mapping and Spatial Analysis with R](https://cengel.github.io/rspatial/)

* [CRAN Task View: Analysis of Spatial Data](https://cran.r-project.org/web/views/Spatial.html)

* [Mapping Spatial Data in R](https://hautahi.com/rmaps)

* [Raster Data in R - The Basics](https://www.neonscience.org/resources/learning-hub/tutorials/raster-data-r)

* [Drawing raster maps with ggmap](https://cfss.uchicago.edu/notes/raster-maps-with-ggmap/)

* Roger S. Bivand, Edzer J. Pebesma, Virgilio Gómez-Rubio, Applied Spatial Data Analysis with R.
